STAT 672 Final Project:\\ Stochastic Gradient Descent

### 2018-04-18 paper/paper.tex
case study

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{graphicx}
\usepackage[linesnumbered, lined, boxed]{algorithm2e}
\RestyleAlgo{boxruled}

@@

@@

This paper is divided into four sections. The remainder of this
\textbf{Introduction} section gives intuitive motivation for stochastic gradient
descent (SGD). The \textbf{Method and Theory} section more rigorously presents the
mathematics of SGD and some of its notable properties. The
\textbf{Applications} sections highlights the real-world settings and uses of SGD,
including a case study data analysis. The \textbf{Conclusion} section summarizes
overall findings.

@@

SGD is commonly used in large-$n$ applications. Thus,
even though SGD is a computational improvement over batch GD, there has been
interest in whether SGD can be made even faster by parallelizing it. As
background, typical computer applications are serial: problems are broken
down into a discrete series of instructions that are executed one after another
on a single processor. Parallel computing calls for breaking down problems into discrete parts that can
be solved concurrently and are executed simultaneously on multiple processors
(\cite{barney2010introduction}).
There are many paradigms of parallel computing but a useful high-level
distinction is between shared memory, in which all processors share a common
address space, and distributed memory, in which each processor or group of processors
has its own address space. The latter has become increasingly popular in
industry via Apache Hadoop and associated projects (e.g. Spark). Distributed
memory applications often require very complex computing strategies for how to combine the
work of many processors without a shared memory space into a single coherent answer.

This background helps to emphasize the novelty of the parallel
implementation of SGD proposed in \cite{zinkevich2010parallelized}. Though backed
by highly technical proofs, the actual algorithm is strikingly simple. Suppose we have $N$ processors. We seek to
conduct SGD to estimate $w$ across $n$ observations with fixed learning rate
$\gamma$ for some fixed number of steps $T$.

@@

Simulation provides a useful example of the computational advantages of SGD for
estimating coefficients in a predictive model. This section outlines such a
brief simulation, annotated with Python code.

Suppose we have feature matrix $\bm{X}_{n \times D}$ with $n=10^6$ and $D=1$
and outcome variable $\bm{Y}_i \sim N(\bm{X}_i,1)$.

We seek to estimate coefficients in a
regression equation of the form $y = \beta_0 + \beta_1 X$ by ordinary least
squares. We do so via two
competing methods: solving directly via $\hat{\bm{\beta}} =
(\bm{X}'\bm{X})^{-1}\bm{X}'\bm{Y}$ and iterative estimation via SGD.

We compare the estimated coefficients and run-time.
\subsection{Real-World Applications}

>>>>

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{graphicx}

\usepackage[linesnumbered, lined, boxed]{algorithm2e}
\RestyleAlgo{boxruled}

@@

\usepackage{listings}
\lstset{frame = single,
breakatwhitespace=false,
breaklines=true,
numbersep=5pt,
showspaces=false,
showtabs=false,
tabsize=2,
basicstyle=\footnotesize}

@@

This paper is divided into four sections. The remainder of this
\textbf{Introduction} section gives intuitive motivation for stochastic gradient
descent (SGD). The \textbf{Method and Theory} section more rigorously derives
SGD, outlines convergence properties, and reviews extensions to the basic
algorithm. The \textbf{Applications} sections highlights SGD's role in
statistical learning, including a simulation analysis to demonstrate SGD's utility
in large-$n$ settings. The \textbf{Conclusion} section summarizes
overall findings.

@@

SGD is commonly used in large-$n$ applications. Thus,
even though SGD is a computational improvement over batch GD, there has been
interest in whether SGD can be made even faster by parallelizing it.
\cite{zinkevich2010parallelized} present a novel technique for doing so. Though backed
by highly technical proofs, the actual algorithm is strikingly simple. Suppose we have $N$ processors. We seek to
conduct SGD to estimate $w$ across $n$ observations with fixed learning rate
$\gamma$ for some fixed number of steps $T$.

@@

Simulation provides a useful example of the computational advantages of SGD. 
Suppose we have one-dimensional feature matrix $\bm{X}$ with $X_i \sim N(0, 1)$ 
and outcome variable $\bm{Y}$ with the
relationship $Y_i \sim (0.5 X_i) + \epsilon$. There are 10 million observations.

\begin{lstlisting}[language=Python]
import numpy as np
X = np.random.randn(int(10e6))
Y = (0.5 * X) + np.randn(1)
\end{lstlisting}

We seek to estimate coefficients in a regression equation of the form $y = \beta_0 + \beta_1 X$ by ordinary least
squares. We do so via two competing methods: batch GD and SGD.
\begin{lstlisting}[language=Python]
def GD(X, Y, gamma, epsilon, maxiter):
	n = Y.size
	w = 1 #initial guess
	for i in range(maxiter):
		Yhat = np.dot(X, w)
		update = w - gamma * (np.dot(X.T, Yhat - Y) / n)
		if (abs(w - update) < epsilon): 
			break
		else:
			w = update
	return(w)
\end{lstlisting}
\begin{lstlisting}[language=Python]
def SGD(X, Y, gamma, epsilon, maxiter):
	n = Y.size
	w = 1 #initial guess
	for i in range(maxiter):
		j = np.random.randint(0, n)
		Yhat = w * X[j]
		update = w - gamma * ((Yhat - Y[j])*X[j])
		if (abs(w - update) < epsilon):
			break
		else:
			w = update
	return(w)
\end{lstlisting}
We compare the estimated coefficients and run-time.
\subsection{Real-World Applications}
